\forestclaw: Hybrid forest-of-octrees AMR for hyperbolic conservation laws

### 2013-07-31 docs/parco13/paper.tex
Edits to the paper

\begin{figure}
\begin{center}
  \includegraphics[angle=-90,width=.7\columnwidth]{pdf/uni_advance_strong-crop}
  \\[1ex]
  \includegraphics[angle=-90,width=.7\columnwidth]{pdf/uni_exchange_strong-crop}
\end{center}
\caption{Strong scaling of MPI parallelism for advection on a
  uniformly refined mesh.  Top: run time of time integration.  Bottom:
  time spent in local and parallel neighbor exchanges.  The number of
  patches is $2^{2 \ell}$ at level $\ell$.  The flat lines for the
  smaller runs are caused by having fewer patches than MPI processes,
  leaving some of the processors idle.  For each data point we use the
  time required by the slowest processor to verify that the load is
  balanced equally.
%  all processes. \comment{This last
%    sentence is a bit unclear.  What do you mean exactly?}
}
\label{fig:uniscale}
\end{figure}%
We begin our analysis of Example 1 with uniformly-refined mesh experiments run on TACC's
Stampede supercomputer at different levels of resolution.
We use 1 to 16 MPI processes on one 16-core compute node, obtaining a strong
scaling efficiency for the time integration between 97\%--103\% depending on
the resolution (from levels 1 and 8).  Then, we examine processor counts from
16 to 256 using 16 cores per node.  The scaling behavior remains linear for
the time integration itself, and mostly linear for the local and parallel
neighbor exchanges which require about 10\% of the time integration run time;
see \Fig{uniscale}.

\begin{figure}
\begin{center}
%  \includegraphics[angle=-90,width=.8\columnwidth]{../figs_misc/omp}
%  \\[1ex]
  \includegraphics[angle=-90,width=.7\columnwidth]{../figs_misc/advance_strong_omp}
\end{center}
\caption{Strong scaling of OpenMP parallelism for advection on a
  uniformly refined mesh:
  %Top: run time of advection Riemann problem kernel in both an optimized inlined interface and a generic function pointer interface via \manyclaw.  Bottom:
  MPI + OpenMp hybrid parallelization of the time integration.
}
\label{fig:ompscale}
\end{figure}%

In \Fig{ompscale}, we scale multi-node parallel calculations.
%and in future
%iterations on accelerators 
We use the \manyclaw set of patch-based PDE solvers.
%with
%\forestclaw.  
\manyclaw provides a generic set of interators for Riemann
problems and related algorithms.

@@

\begin{figure}
  \begin{center}
    \hspace{.1\columnwidth}
    \begin{minipage}[c]{.35\columnwidth}
      \includegraphics[width=.99\columnwidth]{../figs_misc/sphere_partition5}
    \end{minipage}
    \hspace{.05\columnwidth}
    \begin{minipage}[c]{.4\columnwidth}
    \begin{tabular}{r|r}
       $P$ & time int./step \\
      \hline
       16 & 1.115 \\
       32 & 0.557 \\
       64 & 0.297 \\
      256 & 0.092 \\
    \end{tabular}
  \end{minipage}
\end{center}
\caption{Left: Partition of the spherical Example 2 between five MPI processors, indicated
  by the color scale, at simulated time $T = 1$.  The partition shows the mesh equator,
  while the refinement is invariant under the number of processors and the partition
  and follows the front, which is not aligned with the equator.  Right: Strong scaling
  for a uniform run at level 8.}
\label{fig:spherical}
\end{figure}
To demonstrate a multiblock connectivity, we include results for Example 2 in
\Fig{spherical}, together with a strong scaling table.

@@

We have presented a novel hybrid paradigm for AMR that achieves good
scaling both in the distributed and shared memory aspect of current
supercomputers.
%
% By
% abstracting a leaf in the AMR hierarchy as the atomic unit of computation, we
% are able to use almost arbitrary low level solvers.
%
Our implementation integrates an MPI-based forest-of-octrees adaptive
meshing code, \pforest, with a numerical solver for hyperbolic
conservation laws, \clawpack/\manyclaw,
that implements threaded parallelism for a single compute patch.
%All topological operations on the mesh are handled by \pforest,
%while all numerical operations are handled by the \clawpack instance.
This combination naturally enables MPI/threaded hybridization,
where the MPI version scales to 16,384 CPU cores.

@@

We would like to thank the Texas Advanced Computing Center (TACC) for access to
the Stampede supercomputer under allocation TG-DPP130002 and TG-ASC130001 granted by the NSF XSEDE Program. 
The authors acknowledge valuable discussion with Randy LeVeque, Marsha
Berger, and Hans-Petter Langtangen.  We also acknowledge David
Ketcheson and the KAUST sponsored HPC$^3$ numerics workshop at which
the initial phases of this project were first discussed.  The second
author would like to also acknowledge the Isaac Newton Institute
(Cambridge, UK), where much of the preliminary development work for
\forestclaw was done.
The fourth author recognizes Simula National Lab for funding.
The leaf/patch paradigm was independently
presented by B.\ as part of a talk at the SCI Institute, Salt Lake City, Utah,
USA in July 2011.

>>>>

\begin{figure}
\begin{center}
  \includegraphics[angle=-90,width=.75\columnwidth]{pdf/uni_advance_strong-crop}
  \\[1ex]
  \includegraphics[angle=-90,width=.75\columnwidth]{pdf/uni_exchange_strong-crop}
\end{center}
\caption{Strong scaling of MPI parallelism for advection on a
  uniformly refined mesh.  Top: run time of time integration.  Bottom:
  time spent in local and parallel neighbor exchanges.  The number of
  patches is $2^{2 \ell}$ at level $\ell$.  The flat lines for the
  smaller runs are caused by having fewer patches than MPI processes,
  leaving some of the processors idle.  For each data point we use the
  time required by the slowest processor to verify that the load is
  balanced equally.
%  all processes. \comment{This last
%    sentence is a bit unclear.  What do you mean exactly?}
}
\label{fig:uniscale}
\end{figure}%
We begin our analysis of Example 1 with uniformly-refined mesh experiments run on TACC's
Stampede supercomputer at different levels of resolution.
We use 1 to 16 MPI processes on one 16-core compute node, obtaining a strong
scaling efficiency for the time integration between 97\%--103\% depending on
the resolution (from levels 1 and 8).  Then, we examine processor counts from
16 to 256 using 16 cores per node.  The scaling behavior remains linear for
the time integration itself, and mostly linear for the local and parallel
neighbor exchanges which require about 10\% of the time integration run time;
see \Fig{uniscale}.

\begin{figure}
\begin{center}
%  \includegraphics[angle=-90,width=.8\columnwidth]{../figs_misc/omp}
%  \\[1ex]
  \includegraphics[angle=-90,width=.75\columnwidth]{../figs_misc/advance_strong_omp}
\end{center}
\caption{Strong scaling of MPI/OpenMP parallelism for advection on a
  uniformly refined mesh with 16 threads per node.
  %Top: run time of advection Riemann problem kernel in both an optimized
  %inlined interface and a generic function pointer interface via \manyclaw.
  %Bottom:
  %MPI + OpenMp hybrid parallelization of the time integration.
}
\label{fig:ompscale}
\end{figure}%

In \Fig{ompscale}, we show multi-node calculations using 16 OpenMP
threads per node executed by the
%and in future
%iterations on accelerators 
\manyclaw set of patch-based PDE solvers.
%with
%\forestclaw.  
%\manyclaw provides a generic set of interators for Riemann
%problems and related algorithms.

@@

\begin{figure}
  \begin{center}
    \hspace{.1\columnwidth}
    \begin{minipage}[c]{.35\columnwidth}
      \includegraphics[width=.99\columnwidth]{../figs_misc/sphere_partition5}
    \end{minipage}
    \hspace{.05\columnwidth}
    \begin{minipage}[c]{.4\columnwidth}
    \begin{tabular}{r|r}
       $P$ & time int./step \\
      \hline
       16 & 1.115 \\
       32 & 0.557 \\
       64 & 0.297 \\
      256 & 0.092 \\
    \end{tabular}
  \end{minipage}
\end{center}
\caption{Left: Partition of the spherical Example 2 between five MPI processors, indicated
  by the color scale, at simulated time $T = 1$.  The partition shows the mesh equator,
  while the refinement is invariant under the number of processors and the partition
  and follows the front, which is not aligned with the equator.  Right: Strong scaling
  for a uniform run at level 8.}
\label{fig:spherical}
\end{figure}
To showcase a multiblock connectivity, we include results for the spherical Example 2 in
\Fig{spherical}, together with a strong scaling table from 16 to 256 MPI ranks.

@@

We have presented a novel hybrid paradigm for AMR that achieves good
scaling both in the distributed and shared memory aspects of current
supercomputers.
%
% By
% abstracting a leaf in the AMR hierarchy as the atomic unit of computation, we
% are able to use almost arbitrary low level solvers.
%
Our implementation integrates an MPI-based forest-of-octrees adaptive
meshing code, \pforest, with a numerical solver for hyperbolic
conservation laws, \clawpack/\manyclaw,
that implements threaded parallelism for a single compute patch.
%All topological operations on the mesh are handled by \pforest,
%while all numerical operations are handled by the \clawpack instance.
This combination naturally enables MPI/threaded hybridization,
where our biggest MPI run scales to 16,384 CPU cores.

@@

We would like to thank the Texas Advanced Computing Center (TACC) for access to
the Stampede supercomputer under allocations TG-DPP130002 and TG-ASC130001
granted by the NSF XSEDE Program.
The authors acknowledge valuable discussion with Randy LeVeque, Marsha
Berger, and Hans-Petter Langtangen.  We also acknowledge David
Ketcheson and the KAUST sponsored HPC$^3$ numerics workshop at which
the initial phases of this project were first discussed.  The second
author would like to also acknowledge the Isaac Newton Institute
(Cambridge, UK), where much of the preliminary development work for
\forestclaw was done.
The fourth author recognizes Simula National Lab for funding.
The leaf/patch paradigm was independently
presented by B.\ as part of a talk at the SCI Institute, Salt Lake City, Utah,
USA in July 2011.
